---
title: "Analyzing trends in Canadian glacier mass"
author: "Khodosevich Leonid"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

The data used in this study is taken from <http://open.canada.ca/en/open-data>.

```{r, echo=FALSE}
glacier <- read.csv("glacier.csv", na.strings = "..", comment.char = "#")
glacier$GEO = factor(glacier$GEO)
```

The data set contains `r nrow(glacier)` measurements of `r nlevels(glacier$GEO)` Canadian glacier mass balance, collected from `r min(glacier$Ref_Date)` till `r max(glacier$Ref_Date)`. Namely, the file includes these glaciers:

```{r, echo=FALSE}
levels(glacier$GEO)
```

```{r, echo=FALSE}
library(dplyr)
glacier %>%  group_by(GEO) %>% summarise(n = n()) %>%  arrange(n) %>% filter(row_number()==1) %>% select(GEO)
glacier %>%  group_by(GEO, MEASURE) %>% summarise(med = median(Value, na.rm = T)) %>% arrange(desc(med)) %>% head(1) %>% select(GEO)
glacier$GEO[is.na(glacier$Value)]
```

## Excercise
Among the six glaciers studied, the one with the shortest recorded history has the following 
The glacier for which the median value of change (the Value variable for Annual Mass Balance) is closest to zero is called the : The only missing value contains the observation history of the glacier

